---
title: "OrkambiKIDS Statistcal Analysis"
author: "Virginia Rossow"
date: "20 6 2022"
output: html_document
---

```{r setup2, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r path and packages2, include=FALSE}

library(pacman)
pacman::p_load(tidyverse, magrittr, janitor, microbiome, knitr, naniar, phyloseq, LongDat, kableExtra,
               textshape, Biostrings, markdown, vegan, rstatix, gdata, tableone, corrplot, PerformanceAnalytics)

ps_orkambi_work <- readRDS("ps_orkambi_trimmed.rds")

```


```{r run scripts, include=FALSE}

# # run script to visualize the data
# rmarkdown::render("orkambi_kids_visualizations.Rmd", "html_document")

```

      

```{r prepare cinical_table, include=FALSE}

clinical_table_1_9 <- as(sample_data(ps_orkambi_work),"data.frame")

#clinical_table_1_9 <- clinical_table_1_9[!(row.names(clinical_table_1_9) %in% c("19SAV1St", "19MLV5St")), ]

clinical_table_1_9 <- clinical_table_1_9[!duplicated(clinical_table_1_9[ , c("probe_id")]), ]

```

```{r cohort characteristics}

cohort_df <- clinical_table_1_9 %>%
   select(c(visit, probe_id, age, pp_fev1, pp_fvc, bmi_perz, lci)) %>%
   filter(visit == 1 | visit == 9 | visit ==5)

cohort_df <- cohort_df[!(row.names(cohort_df) %in% c("19SAV1St", "19MLV5St")), ]

cohort_df <- cohort_df[!duplicated(cohort_df[ , c("probe_id")]), ]

#data.frame(sapply(cohort_df %>% filter(visit == 1) %>% select(-c(probe_id, visit)), quantile))

IQR((cohort_df %>% filter(visit == 5))$bmi_perz, na.rm = TRUE)


```



## Basic tests on the Metadata

```{r basic exploration, warning=FALSE}

# Events per variable
(nrow(clinical_table_1_9)/7) %>% floor # -> max. 8 variables for per model
(nrow(clinical_table_1_9)/15) %>% floor # -> max. 4 variables for per model (conservative)

summary(clinical_table_1_9$pp_fev1) # normal distribution possible, probably

clinical_num <- clinical_table_1_9 %>% 
   dplyr::select_if(is.numeric) %>%
   select(-c(kreon_average_i_eper_meal, pankreasenzyme, x1_calp_messung,
             bsg_1h_mm, v1bsg_2h_mm, total_leucocytes_10_6, gew_zscore, groe_zscore, bmi_zscore, tn_falpha, il1beta, il6, il8,
             ab90vorher, ab15vorher, ab90burden, ab90treatment, ab90courses, ab15burden, ab15treatment, ab15courses,
             mean_leucocyte_count_10_4_ml, x2_calp_messung, weight_kg, length_cm, bmi, sp_o2_percent, fev1, 
             pred_fev1, fvc, pred_fvc, lci, pred_lci, ig_a, ig_m, ab365vorher, frc, cathepsin))
# cathepsin - outlier because of one patient
# everything with >60% NA

vis_miss(clinical_num)

# chart.Correlation(clinical_num %>% select(-c(gew_perz, groe_perz, ab365burden, pred_frc, crp, ab356courses)),
#                   method = "pearson",
#                   histogram = TRUE,
#                   pch = 16)


```

## Linear Models

ID als random effect

```{r simple Regressions and mixed models}

hist(clinical_table_1_9$pp_fev1, breaks = 12)
clinical_table_1_9 %>% filter(pp_fev1 <= 70) # make NA

clinical_table_1_9 <- clinical_table_1_9 %>%
   replace_with_na_at(.vars = c("pp_fev1"), condition = ~.x <= 70)


lmerTest::lmer(pp_fev1 ~ visit + (1|id), clinical_table_1_9) %>% summary()

lmerTest::lmer(lci ~ visit + (1|id), clinical_table_1_9) %>% summary()
lmerTest::lmer(lci ~ visit + (1|id), clinical_table_1_9 %>% filter(visit != 6, visit != 7, visit != 8, visit != 9)) %>% summary()

pairwise.wilcox.test(clinical_table_1_9$lci, clinical_table_1_9$visit,  p.adjust.method = "BH")

ggplot(clinical_table_1_9, aes(x = as.numeric(visit), y = lci)) +
   geom_jitter() +
   geom_smooth(method = lm)

#median((clinical_table_1_9 %>% filter(visit == 5))$sweatchloride_mmol_l)

lmerTest::lmer(sweatchloride_mmol_l ~ visit + (1|id), clinical_table_1_9 %>% filter(visit == 1 | visit == 2 | visit == 5)) %>% summary() # nope

lmerTest::lmer(pp_fev1 ~ bmi_perz + (1|id), clinical_table_1_9) %>% summary()

ggplot(clinical_table_1_9, aes(x = bmi_perz, y = pp_fev1)) +
   geom_jitter() +
   geom_smooth(method = lm)

hist(clinical_table_1_9$ig_g, breaks = 12)

lmerTest::lmer(pp_fev1 ~ ig_g + (1|id), clinical_table_1_9) %>% summary() # yes

ggplot(clinical_table_1_9, aes(x = ig_g, y = pp_fev1)) +
   geom_jitter() +
   geom_smooth(method = lm)

hist(clinical_table_1_9$leukos, breaks = 12)

lmerTest::lmer(leukos ~ visit + (1|id), clinical_table_1_9) %>% summary() # yes, at some points

ggplot(clinical_table_1_9, aes(x = as.numeric(visit), y = leukos)) +
   geom_point() +
   geom_smooth(method = "glm")

hist(clinical_table_1_9$ab90vorher, breaks = 12)

lmerTest::lmer(ab90vorher ~ visit + (1|id), clinical_table_1_9) %>% summary() # no

ggplot(clinical_table_1_9, aes(x = as.numeric(visit), y = ab90vorher)) +
   geom_jitter() +
   geom_smooth(method = "glm")

hist(clinical_table_1_9$bmi_perz, breaks = 12)

lmerTest::lmer(bmi_perz ~ visit + (1|id), clinical_table_1_9) %>% summary() # visit 7

ggplot(clinical_table_1_9, aes(x = as.numeric(visit), y = bmi_perz)) +
   geom_jitter() +
   geom_smooth(method = "glm")


```


```{r multiple Regressions}

# Events per variable
(nrow(clinical_table_1_9)/15) %>% floor # -> max. 8 variables for per model (conservative)
(nrow(clinical_table_1_9)/7) %>% floor # -> max. 4 variables for per model (not so conservative)

# Models
lmerTest::lmer(pp_fev1 ~ bmi_perz + ig_g + leukos + (1|id), clinical_table_1_9) %>% summary() # only ig_g

lmerTest::lmer(pp_fev1 ~ bmi_perz*visit + (1|id), clinical_table_1_9) %>% summary() # nope


```

### Investigate the stool issue

```{r stool tests, include=FALSE}

# # shannon, observed and simpson individually over the visists
# 
# # richness for every single sample, shannon
# rich_shannon <- estimate_richness(ps_orkambi_stool, split = TRUE, measures = c("Shannon"))
# #view(rich_shannon)
# 
# rich_shannon_tbl <- dplyr::as_tibble(rich_shannon, rownames = "samp_id")
# 
# # make new column with visit
# rich_shannon_tbl <- rich_shannon_tbl %>%
#    mutate(visit = str_sub(samp_id, 6, 6)) %>%
#    mutate(visit = as.numeric(visit)) %>%
#    mutate(visit = as_factor(visit)) %>%
#   clean_names()
# 
# lm(shannon ~ visit, data = rich_shannon_tbl) %>% summary()
# 
# # richness for every single sample, simpson
# rich_simpson <- estimate_richness(ps_orkambi_stool, split = TRUE, measures = c("Simpson"))
# 
# rich_simpson_tbl <- dplyr::as_tibble(rich_simpson, rownames = "samp_id")
# 
# # make new column with visit
# rich_simpson_tbl <- rich_simpson_tbl %>%
#    mutate(visit = str_sub(samp_id, 6, 6)) %>%
#    mutate(visit = as.numeric(visit)) %>%
#    mutate(visit = as_factor(visit)) %>%
#   clean_names()
# 
# lm(simpson ~ visit, data = rich_simpson_tbl) %>% summary()


```

## Permanovas

```{r permanovas}

ps_orkambi_1_9_rel <- transform_sample_counts(ps_orkambi_work, function(x) x/sum(x))

ps_orkambi_1_9_rel_sp <- subset_samples(ps_orkambi_1_9_rel, material== "Sputum")
ps_orkambi_1_9_rel_th <- subset_samples(ps_orkambi_1_9_rel, material== "Throat")
ps_orkambi_1_9_rel_st <- subset_samples(ps_orkambi_1_9_rel, material== "Stool")

# Sputum

metadata_sputum <- as(sample_data(ps_orkambi_1_9_rel_sp),"data.frame")
features_sputum <- as.data.frame(otu_table(ps_orkambi_1_9_rel_sp))

vegan::adonis2(features_sputum ~ visit + bmi_perz + id + pp_fev1,data=metadata_sputum, permutations = 999, na.action=na.exclude, method = "bray") %>% tidy()

# Throat

metadata_throat <- as(sample_data(ps_orkambi_1_9_rel_th),"data.frame")
features_throat <- as.data.frame(otu_table(ps_orkambi_1_9_rel_th))

vegan::adonis2(features_throat ~ visit +bmi_perz + id + pp_fev1,data=metadata_throat, permutations = 999, na.action=na.exclude, method = "bray") %>% tidy()

# Stool

metadata_stool <- as(sample_data(ps_orkambi_1_9_rel_st),"data.frame")
features_stool <- as.data.frame(otu_table(ps_orkambi_1_9_rel_st))

vegan::adonis2(features_stool ~ visit +bmi_perz + id + pp_fev1,data=metadata_stool, permutations = 999, na.action=na.exclude, method = "bray") %>% tidy()


```


## MetaDeconfoundR


```{r metadconfoundr, include=FALSE, eval=FALSE}

# install.packages("remotes")
# remotes::install_github("TillBirkner/metadeconfoundR")
library(metadeconfoundR)

ps_orkambi_work <- readRDS("/fast/AG_Forslund/Rebecca/CF_Scripts/ps_orkambi_trimmed.rds")

# general on genus level

ps_orkambi_metad_glom <- tax_glom(ps_orkambi_work, "genus")

metadata_orkambi_metad <- as(sample_data(ps_orkambi_metad_glom),"data.frame")

#metadata_orkambi_metad %>% filter(patient_id == "SA11")

# remove the unnecessary variables
metadata_orkambi_metad <- metadata_orkambi_metad %>%
   select(-c(visit_date, be_date, throat_swab_date, stool_date, urine_date, weight_kg, length_cm, bmi, start_orkambi, sp_o2_percent, fev1, pred_fev1, fvc, pred_fvc, pred_lci, pp_lci, frc, pred_frc, total_leucocytes_10_6, squamous_cell_count_percent, filtratweight_g, filtratvolume_ml, kreon_average_i_eper_meal, gew_zscore, groe_zscore, bmi_zscore, sample_name, target_name, amp_status, task, intercept, r_squared, slope, efficiency, auto_threshold, auto_baseline, baseline_start, omit, date_q_pcr, extraction_date, project, kit, amount_stool_mg, material_label))

# make all factors numeric 
# -> dadurch gehen alle eins hoch, insbesondere die Mikrobio und Medikamente. Ist das schlimm?
metadata_orkambi_metad <- metadata_orkambi_metad %>%
   mutate(across(id:patient_id, as.numeric)) %>%
   mutate(across(visit:age, as.numeric))

metadata_orkambi_metad <- metadata_orkambi_metad %>%
   mutate(gender = (gender-1)) %>%
   mutate(base = (base - 1))

# select useful variables (by omitting useless ones)
metadata_orkambi_metad <- metadata_orkambi_metad %>%
   select(-c(patient_id, probe_id, mean_leucocyte_count_10_4_ml, salbutamol, cortison_nasenspray, cortison_inhal, dexamethason_infectodexakrupp_b_b, aspecton_nasenspray, vitamines, pancreatic_enzymes, naproxen, luffanest_homeopath, ab365vorher, ab90vorher, ab15vorher, ab365burden, ab90burden, ab15burden, cq_mean, cq_standard_deviation, quantity_mean, quantity_standard_deviation, unnamed_6, position, extract_quant_ng_ul, grouping)) 
# grouping zeigt die a proben nicht als extra an -> raus

# binary variable at the front i guess
metadata_orkambi_metad <- metadata_orkambi_metad %>%
   relocate(base, .before = id)

# igg_response
# 0 = bad
# 1 = good
metadata_orkambi_metad <- metadata_orkambi_metad %>%
   mutate(igg_response = case_when(id == 2 | id == 3 | id == 4 ~ 0,
                                 TRUE ~ 1)) %>%
   relocate(igg_response, .before = base)

# remove sequencing metadata
metadata_orkambi_metad <- metadata_orkambi_metad %>%
   select(-c(well:dna_quant_ng_ul, ul_eluat:interim_sample))


# sputum on genus level

ps_orkambi_metad_glom_sp <- subset_samples(ps_orkambi_metad_glom, material == "Sputum")

# metadata
metadata_orkambi_metad_sp <- metadata_orkambi_metad %>%
   filter(material == "Sputum")

metadata_orkambi_metad_sp$igg_response # at least 5 per group

# other input tables
features_sp <- as.data.frame(otu_table(ps_orkambi_metad_glom_sp))
taxtable_sp <- as.data.frame(tax_table(ps_orkambi_metad_glom_sp))


taxtable_sp <- taxtable_sp %>%
  select(genus)

features_sp_t <- as.data.frame(t(features_sp))


features_sp_named <- features_sp_t
features_sp_named <- merge(features_sp_named, taxtable_sp, by = 0)
row.names(features_sp_named) <- features_sp_named$genus
features_sp_named$Row.names <- NULL
features_sp_named$genus <- NULL
features_sp_named <- as.data.frame(t(features_sp_named))

# metadeconfound sputum
sputum_metad <- MetaDeconfound(featureMat = features_sp_named,metaMat = metadata_orkambi_metad_sp, randomVar = list("+ (1|id)", c("id")), nnodes=6)

# heatmap
raw_p <- sputum_metad[1]
corr_p <- sputum_metad[2]
effect_size <- sputum_metad[3]
status <- sputum_metad[4]

hm <- BuildHeatmap(sputum_metad, showConfounded = F, q_cutoff = 0.05, d_cutoff = 0.01) 
hm +
  theme(axis.text.x=element_text(size=16, angle = 40, vjust = 1), axis.text.y=element_text(size=16, face="italic"), legend.position = "bottom")+
  xlab("")+
  ylab("")

ggsave("/fast/AG_Forslund/Rebecca/CF_Scripts/heatmapMetadeconfoundR_sp_f_igg.png", width = 23, height = 17)


# throat on genus level

ps_orkambi_metad_glom_th <- subset_samples(ps_orkambi_metad_glom, material == "Throat")

# metadata
metadata_orkambi_metad_th <- metadata_orkambi_metad %>%
   filter(material == "Throat")

metadata_orkambi_metad_th$igg_response

# other input tables
features_th <- as.data.frame(otu_table(ps_orkambi_metad_glom_th))
taxtable_th <- as.data.frame(tax_table(ps_orkambi_metad_glom_th))

taxtable_th <- taxtable_th %>%
  select(genus)

features_th_t <- as.data.frame(t(features_th))

features_th_named <- features_th_t
features_th_named <- merge(features_th_named, taxtable_th, by = 0)
row.names(features_th_named) <- features_th_named$genus
features_th_named$Row.names <- NULL
features_th_named$genus <- NULL
features_th_named <- as.data.frame(t(features_th_named))

# metadeconfound throat
throat_metad <- MetaDeconfound(featureMat = features_th_named,metaMat = metadata_orkambi_metad_th, randomVar = list("+ (1|id)", c("id")), nnodes=6)

# heatmap
raw_p <- throat_metad[1]
corr_p <- throat_metad[2]
effect_size <- throat_metad[3]
status <- throat_metad[4]

hm <- BuildHeatmap(throat_metad, showConfounded = F, q_cutoff = 0.05, d_cutoff = 0.01) 
hm +
  theme(axis.text.x=element_text(size=16, angle = 40, vjust = 1), axis.text.y=element_text(size=16, face="italic"), legend.position = "bottom")+
  xlab("")+
  ylab("")

ggsave("/fast/AG_Forslund/Rebecca/CF_Scripts/heatmapMetadeconfoundR_th_f_igg.png", width = 23, height = 17)


# stool on genus level

ps_orkambi_metad_glom_st <- subset_samples(ps_orkambi_metad_glom, material == "Stool")

# metadata
metadata_orkambi_metad_st <- metadata_orkambi_metad %>%
   filter(material == "Stool")

metadata_orkambi_metad_st$igg_response

# other input tables
features_st <- as.data.frame(otu_table(ps_orkambi_metad_glom_st))
taxtable_st <- as.data.frame(tax_table(ps_orkambi_metad_glom_st))

taxtable_st <- taxtable_st %>%
  select(genus)

features_st_t <- as.data.frame(t(features_st))

features_st_named <- features_st_t
features_st_named <- merge(features_st_named, taxtable_st, by = 0)
row.names(features_st_named) <- features_st_named$genus
features_st_named$Row.names <- NULL
features_st_named$genus <- NULL
features_st_named <- as.data.frame(t(features_st_named))

# metadeconfound stool
stool_metad <- MetaDeconfound(featureMat = features_st_named,metaMat = metadata_orkambi_metad_st, randomVar = list("+ (1|id)", c("id")), nnodes=6)

# heatmap
raw_p <- stool_metad[1]
corr_p <- stool_metad[2]
effect_size <- stool_metad[3]
status <- stool_metad[4]

hm <- BuildHeatmap(stool_metad, showConfounded = F, q_cutoff = 0.05, d_cutoff = 0.01) 
hm +
  theme(axis.text.x=element_text(size=16, angle = 40, vjust = 1), axis.text.y=element_text(size=16, face="italic"), legend.position = "bottom")+
  xlab("")+
  ylab("")

ggsave("/fast/AG_Forslund/Rebecca/CF_Scripts/heatmapMetadeconfoundR_st_f_igg.png", width = 23, height = 17)

```


## Side effects

```{r side effects}

side_effects_tbl <- clinical_table_1_9 %>%
   select(-contains("date")) %>%
   select(-c(start_orkambi)) %>%
   select(-c(84:113)) 

# select side effects which occurred most often

sides <- side_effects_tbl %>%
   select(c(probe_id,59:73)) %>%
   mutate(across(where(is.factor), as.numeric))
sides <- sides[!duplicated(sides[ , c("probe_id")]), ] 
sides <- sides[!(row.names(sides) %in% c("19MLV5St")), ]
rownames(sides) <- sides$probe_id
sides$probe_id <- NULL

sides[is.na(sides)] <- 1

sides_1 <- (sides -1)

table(sides_1$ubelkeit)

sides_occ <- c(19, 10, 29, 29, 13, 20, 3, 10, 6, 36, 20, 43, 14, 1, 13)
sides_occ_df <- data.frame(side_effects = colnames(sides_1), side_effects_occ = sides_occ)

ggplot(sides_occ_df, aes(x = side_effects, y = side_effects_occ)) +
   theme_bw() +
   geom_point() +
   #geom_segment( aes(x=x, xend=x, y=0, yend=y)) +
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
   xlab("side effects") +
   ylab("number of occurences")

sides_1 %>% colSums()

blahungen_tbl <- side_effects_tbl %>% filter(!is.na(blahungen))

table(side_effects_tbl$blahungen, side_effects_tbl$visit)

ggplot(blahungen_tbl, aes(x = visit, y = blahungen, color = blahungen)) +
   geom_jitter() 


# Blähungen
barplot(table(side_effects_tbl$blahungen)) %>%
   title(main = "Blähungen", ylab = "Häufigkeit", xlab = "Stärke")

# Bauchschmerzen
barplot(table(side_effects_tbl$bauchschmerzen)) %>%
   title(main = "Bauchschmerzen", ylab = "Häufigkeit", xlab = "Stärke")

# laufende Nase
barplot(table(side_effects_tbl$laufende_nase)) %>%
   title(main = "Laufende Nase", ylab = "Häufigkeit", xlab = "Stärke")

# verstopfte Nase
barplot(table(side_effects_tbl$verstopfte_nase)) %>%
   title(main = "Verstopfte Nase", ylab = "Häufigkeit", xlab = "Stärke")

# vermehrter Husten
barplot(table(side_effects_tbl$vermehrter_husten)) %>%
   title(main = "Vermehrter Husten", ylab = "Häufigkeit", xlab = "Stärke")

# Kopfschmerzen
barplot(table(side_effects_tbl$kopfschmerzen)) %>%
   title(main = "Kopfschmerzen", ylab = "Häufigkeit", xlab = "Stärke")

# hier dann die eintwicklung der häufigsten Nebenwirkungen über die Zeit...

par(mfrow=c(2,2))
barplot(table(side_effects_tbl %>% filter(visit == 1) %>% select(blahungen))) %>%
   title(main = "Blähungen - Visit 1", ylab = "Häufigkeit", xlab = "Stärke")
barplot(table(side_effects_tbl %>% filter(visit == 2) %>% select(blahungen))) %>%
   title(main = "Blähungen - Visit 2", ylab = "Häufigkeit", xlab = "Stärke")
barplot(table(side_effects_tbl %>% filter(visit == 5) %>% select(blahungen))) %>%
   title(main = "Blähungen - Visit 5", ylab = "Häufigkeit", xlab = "Stärke")
barplot(table(side_effects_tbl %>% filter(visit == 9) %>% select(blahungen))) %>%
   title(main = "Blähungen - Visit 9", ylab = "Häufigkeit", xlab = "Stärke")


par(mfrow=c(3,3))
barplot(table(side_effects_tbl %>% filter(visit == 1) %>% select(blahungen))) %>%
   title(main = "Blähungen - Visit 1", ylab = "Häufigkeit", xlab = "Stärke")
barplot(table(side_effects_tbl %>% filter(visit == 2) %>% select(blahungen))) %>%
   title(main = "Blähungen - Visit 2", ylab = "Häufigkeit", xlab = "Stärke")
barplot(table(side_effects_tbl %>% filter(visit == 3) %>% select(blahungen))) %>%
   title(main = "Blähungen - Visit 3", ylab = "Häufigkeit", xlab = "Stärke")
barplot(table(side_effects_tbl %>% filter(visit == 4) %>% select(blahungen))) %>%
   title(main = "Blähungen - Visit 4", ylab = "Häufigkeit", xlab = "Stärke")
barplot(table(side_effects_tbl %>% filter(visit == 5) %>% select(blahungen))) %>%
   title(main = "Blähungen - Visit 5", ylab = "Häufigkeit", xlab = "Stärke")
barplot(table(side_effects_tbl %>% filter(visit == 6) %>% select(blahungen))) %>%
   title(main = "Blähungen - Visit 6", ylab = "Häufigkeit", xlab = "Stärke")
barplot(table(side_effects_tbl %>% filter(visit == 7) %>% select(blahungen))) %>%
   title(main = "Blähungen - Visit 7", ylab = "Häufigkeit", xlab = "Stärke")
barplot(table(side_effects_tbl %>% filter(visit == 8) %>% select(blahungen))) %>%
   title(main = "Blähungen - Visit 8", ylab = "Häufigkeit", xlab = "Stärke")
barplot(table(side_effects_tbl %>% filter(visit == 9) %>% select(blahungen))) %>%
   title(main = "Blähungen - Visit 9", ylab = "Häufigkeit", xlab = "Stärke")




```


### For later

Kofidenzintervalle:
k2 <- t.test(chronicity ~ group, data = pr_data) %>% tidy()
k2

str_c(round(k2$estimate, 2), " [",
      round(k2$conf.low, 2), "; ",
      round(k2$conf.high, 2), "]")
      
VIF:
variable inflation factor, zeigt correlation von Variablen (x untereinander) an
"hoher" VIF kann problematisch sein, Daumenregel - VIF < 5 ist ok
ols_vif_tol(model_log_full)

